---
title: "Lab1"
format: pdf
editor: visual
---

## Lab 1

(not finished)

1.  Plot the ratio of male to female mortality rates over time for ages 10,20,30 and 40 (different color for each age) and change the theme

```{r, echo = FALSE, message = FALSE, warning = FALSE}

library(tidyverse)

# Load data set
dm <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Mx_1x1.txt", skip = 2, col_types = "dcddd")
#head(dm)

ratio_fm <- dm  %>% filter(Age %in% c(10,20,30,40)) %>% mutate(ratio = Male/Female)

ratio_fm %>% ggplot(aes(x = Year, y = ratio, colour = Age)) + geom_line() + theme_bw()


```

2.  Find the age that has the highest female mortality rate each year

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dm %>% group_by(Year) %>% filter(Female == max(Female, na.rm = T)) %>% select(Year, Age)
```

3.  Use the \`summarize(across())\` syntax to calculate the standard deviation of mortality rates by age for the Male, Female and Total populations.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dm  %>% group_by(Age) %>% summarise(across(c(Male, Female, Total), sd, na.rm = T)) 


```

4.  The Canadian HMD also provides population sizes over time (https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt). Use these to calculate the population weighted average mortality rate separately for males and females, for every year. Make a nice line plot showing the result (with meaningful labels/titles) and briefly comment on what you see (1 sentence). Hint: \`left_join\` will probably be useful here.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
dsize <- read_table("https://www.prdh.umontreal.ca/BDLC/data/ont/Population.txt", skip = 2, col_types = "dcddd")

colnames(dsize)[3:5] <- c("PopF", "PopM", "PopTotal")

dtot <- left_join(dm, dsize, by = c("Year", "Age")) %>% mutate(DeathsF = (Female*PopF), DeathsM = (Male*PopM)) %>% group_by(Year) %>% summarise(TD_M = sum(DeathsM, na.rm = T), TD_F = sum(DeathsF, na.rm =T), PopTotal = sum(PopTotal, na.rm = T)) 

dw <- dtot %>% mutate(weighted_f = TD_F/PopTotal, weighted_m = TD_M/PopTotal)

dw %>% pivot_longer(c(weighted_f, weighted_m), names_to = "Sex", values_to = "Weighted") %>% filter(Weighted != Inf) %>% ggplot(aes(x = Year, y = Weighted, colour = Sex)) + geom_line() +labs()
```
